\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-08 paper/ms.tex
cleaning up appendix and removing extended

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
\citet{Ambikasaran:2015} derived a scalable direct solver for any general
semiseparable matrix and applied this method to \celeriteterm\ models with
$b_j = 0$ and $d_j = 0$.
In \app{extended}, we generalize this earlier method to include non-zero $b_j$
and $d_j$.
It turns out, however, that we can derive a higher performance algorithm by
only considering positive semi-definite (PSD) matrices.
In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
PSD.

@@

\section{An alternative solver using the extended matrix
formalism}\sectlabel{extended}

\citet{Rybicki:1995} demonstrated that the inverse of a matrix $K$, with
elements given by \eq{kernel-simple}, could be computed
efficiently by taking advantage of the structure of this covariance function.
\citet{Ambikasaran:2015} generalized this computation to apply to the full
mixture of $J$ terms in \eq{celerite-kernel-complex} and derived an equally
efficient method for computing the determinant of $K$.

\subsection{An example}

To provide some insight for this method, we follow
\citet{Ambikasaran:2015} and start by working through a simple example.
In this case, we assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This is the result of setting $J=1$, $b=0$, and $d=0$ in
\eq{celerite-kernel-complex} and it is the model studied by
\citet{Rybicki:1995}.
To demonstrate that the likelihood of this model can be computed in
$\mathcal{O}(N)$, we write out the full system of
equations that must be solved to apply the inverse of $K$ and compute
the first term of \eq{gp-likelihood}.
In matrix notation, this is
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In \eq{impl-matrix}, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$, and
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
Rewriting the system defined by \eq{algo-first} through \eq{algo-last}
as a matrix equation shows the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
Following \citet{Ambikasaran:2015} we call this the ``extended'' system and
rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ we started
with, it is now sparse with banded structure that can be exploited to solve
the system efficiently.
In particular, sparse solvers are available that can perform a
LU-decomposition of matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can use these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

In the following section we discuss this method more generally, but it's
worth noting a few important facts that can already be seen in this example.
First, the fundamental reason why this matrix $K$ can be solved efficiently is
the following property of exponentials
\begin{eqnarray}
    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
    e^{-c\,(t_3 - t_1)}
\end{eqnarray}
and it is important to note that this property does not extend to other common
covariance functions like the ``exponential-squared'' function
\begin{eqnarray}
    k(\tau) \propto e^{-c\,\tau^2} \quad.
\end{eqnarray}
Second, our derivation of the extended matrix requires that the data
points be monotonically sorted.
This is the motivation for our previous statement that this method can only be
applied in one dimension since neither of these properties will be satisfied
by multidimensional inputs.

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix will still be banded with a bandwidth that scales linearly
with the number of components, $J$.
Second, \citet{Ambikasaran:2015} proved that the absolute value of the
determinant of $K_\mathrm{ext}$ is equal to the absolute value of the
determinant of $K$.
This means we can use this extended matrix formalism to compute the
marginalized likelihood in $\mathcal{O}(N)$ operations.

\subsection{The algorithm}

In this section, we generalize the method from the previous section to the
covariance function given by \eq{celerite-kernel}.
This derivation follows \citet{Ambikasaran:2015} but it includes explicit
treatment of complex parameters, and their complex conjugates.

In the case of the full \celeriteterm\ covariance function
(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
analogy to the $u_n$ and $g_n$ that we introduced in the previous section
\begin{eqnarray}\eqlabel{full-system-first}
\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
\end{eqnarray}
with the boundary conditions
\begin{eqnarray}
    g_{0,j} = 0 \quad, \quad
    h_{0,j} = 0 \quad, \quad
    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
    v_{N,j} = 0
\end{eqnarray}
for all $j$.
Using these variables and some algebra, we find that the following expression
\begin{eqnarray}\eqlabel{full-system-last}
\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
    &=& r_{\bvec{\theta},n}
\end{eqnarray}
is equivalent to the target matrix equation
\begin{eqnarray}
K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
\end{eqnarray}
if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
\eq{full-system-first} through \eq{full-system-last} define a banded matrix
equation in the ``extended'' space
\begin{eqnarray}\eqlabel{kext-full}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=&
    \bvec{r}_{\bvec{\theta},\,\mathrm{ext}} \quad,
\end{eqnarray}
and, as before, this can be used to solve
for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
operations.
\Figure{matrix} shows a pictorial representation of the sparsity pattern of
the extended matrix $K_\mathrm{ext}$.
The extended matrix has the dimension $N_\mathrm{ext} = (4\,J+1)\,N-4\,J$,
with a bandwidth $W = 2\,J+2$ and $N_\mathrm{nz} = (20\,J+1)\,N-28\,J$
non-zero entries.
Given this definition of $K_\mathrm{ext}$, the corresponding extended vectors
$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
as
\begin{eqnarray}
{\bvec{z}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
    h_{N-1,j} & z_N
\end{array}\right)
\end{eqnarray}
and
\begin{eqnarray}
{\bvec{r}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
    0 & \cdots & 0 & r_{\bvec{\theta},N}
\end{array}\right) \quad.
\end{eqnarray}
After constructing the extended matrix (using a compact storage format), the
extended matrix can be factorized using a LU-decomposition\footnote{Even
though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
Cholesky solver cannot be used for increased efficiency.} routine optimized
for band or sparse matrices.
This decomposition can then be used to compute the determinant of $K$, solve
$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
likelihood in \eq{gp-likelihood}.

In practice, it is worth treating terms with $b_j=0$ and $d_j=0$ as a special
case because, for this term $j$, $\psi_{n,\,j}$, $h_{n,\,j}$, and $v_{n,\,j}$
will also be identically zero for all $n$.
Removing these trivial rows from the extended matrix results in factor of two
increase in the computational efficiency for these terms.
In our implementation, we refer to ``real'' terms as those where $b_j=0$ and
$d_j=0$, and the general terms are called ``complex''.

\begin{figure}[tp]
\begin{center}
\includegraphics[width=0.7\textwidth]{figures/matrix.pdf}
\caption{A pictorial representation of the sparse extended matrix
$K_\mathrm{ext}$ (\eqalt{kext-full}) with $N=5$ and $J=2$.
Each colored block corresponds to a non-zero entry in the matrix as described
in the legend.
    \figurelabel{matrix}}
\end{center}
\end{figure}

\subsection{Implementation considerations \& scaling}

The extended system defined in the previous section is sparse, with typically
fewer than a few percent non-zero entries and band structure.
In this section, we empirically investigate the performance and scaling
of three different algorithms for solving this extended system:

\begin{enumerate}

\item \texttt{vanilla}: A simple algorithm for computing the LU decomposition
    for banded matrices using Gaussian elimination \citep[we implemented this
    method following][]{Press:1992, Press:2007},

\item \texttt{lapack}: The general banded LU decomposition implementation from
    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
    \project{BLAS} routines,\footnote{The experiments in this paper use the
    Intel Math Kernel Library (\project{MKL}):
    \url{https://software.intel.com/en-us/intel-mkl}.} and

\item \texttt{sparse}: A general sparse LU solver~--~the \texttt{SparseLU}
    solver from \project{Eigen} \citep{Guennebaud:2010}~--~that exploits the
    sparsity but not the band structure.

\end{enumerate}

The theoretical scaling for a band LU decomposition is $\mathcal{O}(N\,J^3)$
because the dimension of the extended matrix scales as $N\,J$ and the
bandwidth scales with $J$ \citep{Press:1992, Press:2007}.
\citet{Ambikasaran:2015} found an empirical scaling of $\mathcal{O}(N\,J^2)$ that used the sparse LU decomposition implemented in the
\project{SuperLU} package \citep{Demmel:1999}.
We find that, while the \texttt{vanilla} solver scales as expected, the
\texttt{lapack} implementation scales empirically as $\mathcal{O}(N\,J^2)$ and
offers the fastest solves for $J \gtrsim 8$ on all platforms that we tested.

The benchmark experiments shown here were performed on a MacBook Pro with two
2.6~GHz CPUs but we find similar results on a Dell workstation with 16 2.7~GHz
CPUs and running Ubuntu.
\Figure{benchmark} shows how the cost of computing \eq{gp-likelihood} scales
with $N$ and $J$ using the \texttt{vanilla} solver.
As expected theoretically, the scaling is linear in $N$ for all $N$ and cubic
in $J$ for large $J$.
\Figure{benchmark-lapack} and \Figure{benchmark-sparse} are the same plots for
the \texttt{lapack} and \texttt{sparse} solvers respectively.
For each of these optimized solvers, the empirical scaling is
$\mathcal{O}(N\,J^2)$ at the cost of some extra overhead.
Therefore, for $J \lesssim 5$ or $10$ the \texttt{vanilla} solver is more
efficient than the other algorithms.
The real world performance of \celerite\ depends on the specific platform,
hardware, and \project{LAPACK}/\project{BLAS} implementation, but we have
found qualitatively similar results across popular platforms and
state-of-the-art libraries.

@@

\section{Ensuring positive definiteness}\sectlabel{psd}

In the case of $J$ \celeriteterm\ terms, we can check for negative values of the
PSD by solving for the roots of the power spectrum; if there are any real,
positive roots, then the power-spectrum goes negative (or zero), and thus does not
represent a valid kernel. We rewrite the power spectrum, \eqalt{celerite-psd}),
abbreviating with $z = \omega^2$:
\begin{equation}
S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_j\,c_j-b_j\,d_j\\
r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by
$\prod_j \left(z^2+s_jz + t_j\right)$ to find
\begin{equation}
Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
    t_k\right) = 0\quad,
\end{equation}
which is a polynomial with order $2\,(J-1)+1$.
With $J=2$, this yields a cubic equation whose roots can be obtained exactly.

>>>>

The relationship between the \celeriteterm\ model and semiseparable linear
algebra was first recognized by Ambikasaran \citep{Ambikasaran:2015} building
on earlier work by Rybicki \& Press \citep{Rybicki:1995}.
\citet{Ambikasaran:2015} derived a scalable direct solver for any general
semiseparable matrix and applied this method to \celeriteterm\ models with
$b_j = 0$ and $d_j = 0$.
During the preparation of this paper, we generalized this earlier method to
include non-zero $b_j$ and $d_j$, but it turns out that we can derive a higher
performance algorithm by restricting our method to positive semi-definite
(PSD) matrices.
In \app{psd}, we discuss methods for ensuring that a \celeriteterm\ model is
PSD.

@@

\section{Ensuring positive definiteness for celerite models}\sectlabel{psd}

\response{The constraint for $J > 1$ is more complicated so, in most cases, we
require that each term is positive definite using this relationship and use
the fact that a product of sum of positive definite kernels will also be
positive definite \citep{Rasmussen:2006} to show that the full model is
positive.}
However, in the case of $J$ general \celeriteterm\ terms, we can check for
negative values of the PSD by solving for the roots of the power spectrum;
if there are any real, positive roots, then the power-spectrum goes negative
(or zero), and thus does not represent a valid kernel.
We rewrite the power spectrum, \eqalt{celerite-psd}), abbreviating with
$z = \omega^2$:
\begin{equation}
S(\omega)=  \sum_{j=1}^J \frac{q_j\,z + r_j}{z^2+s_j\,z + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_j\,c_j-b_j\,d_j\\
r_j &=& (d_j^2+c_j^2)(b_j\,d_j+a_j\,c_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by
$\prod_j \left(z^2+s_jz + t_j\right)$ to find
\begin{equation}
Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
    t_k\right) = 0\quad,
\end{equation}
which is a polynomial with order $2\,(J-1)+1$.
With $J=2$, this yields a cubic equation whose roots can be obtained exactly.

@@
